---
title: "Cyclistic Bike Analysis using R"
output: html_notebook
---

Here, I am going to use R and analyze the processed table to find meaning information for our project. I have combined, cleaned and processed the table for use.

**Here is to review our business task again**: The primary goal of this analysis is to understand how casual riders (who purchase single-ride or full-day passes) and annual members use Cyclistic bikes differently. Insights from this analysis will inform targeted marketing strategies to convert casual riders into annual members.

## **Install and Import Libraries**

These are the libraries that we are going to use for this project.

```{r}
# # Install packages if they are not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse") 
if (!requireNamespace("lubridate", quietly = TRUE)) install.packages("lubridate") 
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr") 
if (!requireNamespace("scales", quietly = TRUE)) install.packages("scales")
if (!requireNamespace("leaflet", quietly = TRUE)) install.packages("leaflet")
if (!requireNamespace("leaflet.extras", quietly = TRUE)) install.packages("leaflet.extras")


# Load the libraries
library(tidyverse)
library(lubridate)
library(readr)
library(leaflet)
library(leaflet.extras)

```

## Load the Dataset

We're going to upload the CSV file for analysis.

```{r}
# Load the CSV file to the DataFrame
cyclistic_data <- read_csv("../../dataset/cleaned/2023_cyclistic_tripdata.csv", 
  col_types = cols(
    started_at = col_datetime(format = "%Y-%m-%d %H:%M:%S %Z"),
    ended_at = col_datetime(format = "%Y-%m-%d %H:%M:%S %Z"),
    day_of_week = col_integer(),
    ride_length = col_time(format = "%H:%M:%S"),
    .default = col_guess()
)) %>% 
  mutate(
    # Modify the day_of_week column to use day names instead of numbers.
    day_of_week = factor(day_of_week, 
                              levels = 1:7, 
                              labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")),
    # Convert ride_length to seconds and overwrite the original column
    ride_length = as.numeric(ride_length, units = "secs")
    )

# Check the structure of the data and print a few rows of the dataset.
str(cyclistic_data)
head(cyclistic_data)

```

## Create custom functions and variables

We are going to create some variables and functions that we can use going forward in our analysis.

```{r}
# Set pre-defined colors for each type.
color_casual <- "#FF5733" # Orange
color_member <- "#2E86C1" # Blue
color_data_point <- "#FFFFFF" # White
```

```{r}
# Custom function to format numbers with K and M
format_k_m <- function(x) {
  ifelse(x >= 1e6, paste0(x / 1e6, "M"),
         ifelse(x >= 1e3, paste0(x / 1e3, "K"), x))
}
```

```{r}
# Custom function to set base aesthetics
set_base_aes <- function(plot_type, x_var, y_var, fill_var) {
  if (plot_type == "pie") {
    aes(
      x = "", y = !!sym(y_var), 
      fill = !!sym(fill_var), color = !!sym(fill_var)
    )
  } else {
    aes(
      x = !!sym(x_var), y = !!sym(y_var), 
      fill = !!sym(fill_var), color = !!sym(fill_var), 
      group = !!sym(fill_var)
    )
  }
}

# Custom function to set the base ggplot ojects.
set_ggplot_objs <- function(data, base_aes, title, x_label, y_label) {
  ggplot(data, base_aes) + 
    scale_fill_manual(
      values = c("casual" = color_casual, "member" = color_member)) +
    scale_color_manual(
      values = c("casual" = color_casual, "member" = color_member)) +
    scale_y_continuous(labels = format_k_m, limits = c(0, NA)) +
    labs(
      title = title, 
      x = x_label, y = y_label, 
      fill = "Rider Type", color = "Rider Type"
      ) +
    theme_minimal()
}

# Custom function to add geom layer based on plot type.
set_geom_layers <- function(base_plot, plot_type, x_var) {
  if (plot_type == "bar") {
    base_plot + 
      geom_bar(stat = "identity", position = position_dodge())
  } else if (plot_type == "pie") {
    base_plot + 
      geom_bar(stat = "identity", width = 1) +
      coord_polar("y", start = 0) +
      theme_void()
  } else if (plot_type == "line") {
    base_plot + 
      geom_line(size = 1) + 
      geom_point(size = 2)
  }
}

# Custom function to add geom_text
set_geom_text <- function(base_plot, label_var, position = "stack", vjust = 0.5, color = color_data_point, size = 4) {
  base_plot + geom_text(
    aes(label = round(!!sym(label_var), 2)),
    position = if (position == "stack") position_stack(vjust = vjust) else position,
    color = color,
    size = size
  )
}

# Function to customize x-axis for hourly data
customize_x_axis <- function(base_plot, x_var) {
  if (!is.null(x_var) && x_var == "hour") {
    base_plot + 
      scale_x_continuous(breaks = 0:23, labels = paste0(0:23, ":00")) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
  } else if (!is.null(x_var) && x_var == "month") {
    base_plot + 
      scale_x_continuous(breaks = 1:12, labels = month.abb) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
  } else if (!is.null(x_var) && x_var == "day_of_week") {
    base_plot + 
      scale_x_discrete(labels = levels(ride_freq_by_day_of_week$day_of_week))
  } else {
    base_plot
  }
}
```

```{r}
# Custom function to create plots.
create_plot <- function(
    data, 
    x_var = NULL, y_var = NULL, fill_var = NULL, 
    plot_type = "bar",
    title, x_label = NULL, y_label = NULL, 
    label_var = NULL, text_position = "stack", text_vjust = 0.5,
    text_color = color_data_point, text_size = 4) {
  
  base_aes <- set_base_aes(plot_type, x_var, y_var, fill_var)
  base_plot <- set_ggplot_objs(data, base_aes, title, x_label, y_label)
  base_plot <- set_geom_layers(base_plot, plot_type, x_var)
  base_plot <- customize_x_axis(base_plot, x_var)
  
  if (!is.null(label_var)) {
    base_plot <- set_geom_text(base_plot, label_var, position = text_position, vjust = text_vjust, color = text_color, size = text_size)
  }
  
  return(base_plot)
  
}
```

## Exploratory Data Analysis

### **Descriptive Analysis**

Descriptive analysis in the context of bike-sharing data involves summarizing the main characteristics of the dataset to provide a clear understanding of how casual riders and annual members use the Cyclistic bikes.

The objective of conducting a descriptive analysis in this project is to establish a foundational understanding of the data at hand.

#### Overall Usage Statistics

```{r}
# Find the average ride_length and distance by rider type.
usage_stats <- cyclistic_data %>% 
  group_by(member_casual) %>% 
  summarise(
    total_rides = n(),
    average_ride_length = mean(ride_length),
    average_distance = mean(trip_distance)
  )  %>%
  # Visualize the percentage of total rides for each rider type.
  mutate(percentage = total_rides / sum(total_rides) * 100,
         avg_ride_min = average_ride_length / 60)

# Visualize a pie chart with total rides as percentage labels
total_rides_plot <- create_plot(
  data = usage_stats,
  x_var = NULL, y_var = "total_rides", fill_var = "member_casual",
  plot_type = "pie",
  title = "Total Rides by Rider Type",
  label_var = "percentage"
)

# Visualize average ride length for each rider type.
average_ride_length_plot <- create_plot(
  data = usage_stats,
  x_var = "member_casual", y_var = "avg_ride_min", fill_var = "member_casual",
  plot_type = "bar",
  title = "Average Ride Length by Rider Type",
  x_label = "Rider Type", y_label = "Average Ride Length (minutes)",
  label_var = "avg_ride_min"
)

# Visualize average trip distance for each rider type.
average_distance_plot <- create_plot(
  data = usage_stats,
  x_var = "member_casual", y_var = "average_distance", fill_var = "member_casual",
  plot_type = "bar",
  title = "Average Distance by Rider Type",
  x_label = "Rider Type", y_label = "Average Distance",
  label_var = "average_distance"
)

# Print all plots.
print(total_rides_plot)
print(average_ride_length_plot)
print(average_distance_plot)
```

### **Temporal Analysis**

#### Ride Frequency by Month

```{r}
# Find the ride frequency by day of week by rider type.
ride_freq_by_month <- cyclistic_data %>% 
  group_by(member_casual, month = month(started_at)) %>% 
  summarise(
    total_rides = n(),
    .groups = "drop"
  )

# Visualize ride frequency by day of week.
ride_freq_by_month_plot <- create_plot(
  data = ride_freq_by_month,
  x_var = "month", y_var ="total_rides", fill_var = "member_casual", 
  plot_type = "line",
  title = "Ride Frequency by Month", 
  x_label = "Month", y_label = "Total Rides"
)

print(ride_freq_by_month_plot)
```

#### Ride Frequency by Season

```{r}
# Find the ride frequency by season by rider type.
ride_freq_by_season <- cyclistic_data %>% 
  mutate(season = factor(
  season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>%
  group_by(member_casual, season) %>% 
  summarise(total_rides = n(),
            .groups = "drop")

# Visualize ride frequency by season.
ride_freq_by_season_plot <- create_plot(
  data = ride_freq_by_season,
  x_var = "season", y_var ="total_rides", fill_var = "member_casual", 
  plot_type = "bar",
  title = "Ride Frequency by Season", 
  x_label = "Season", y_label = "Total Rides"
)

print(ride_freq_by_season_plot)
```

**Ride Frequency by Day of the Week**

```{r}
# Find the ride frequency by day of week by rider type.
ride_freq_by_day_of_week <- cyclistic_data %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(
    total_rides = n(),
    .groups = "drop"
  )

# Visualize ride frequency by day of week.
ride_freq_by_day_of_week_plot <- create_plot(
  data = ride_freq_by_day_of_week,
  x_var = "day_of_week", y_var = "total_rides", fill_var = "member_casual", 
  plot_type = "line",
  title = "Ride Frequency by Day of the Week", 
  x_label = "Day of Week", y_label = "Total Rides"
)

print(ride_freq_by_day_of_week_plot)
```

#### Ride Frequency (Weekend vs. Weekday)

```{r}
# Find the ride frequency on weekends and weekdays by rider type.
ride_freq_by_weekend_weekday <- cyclistic_data %>% 
    mutate(weekday_weekend = if_else(
      is_weekend == "Yes", "Weekend", "Weekday")) %>%
  group_by(member_casual, weekday_weekend) %>% 
  summarise(total_rides = n(),
            .groups = 'drop')

# Visualize ride frequency by weekdays vs weekends.
ride_freq_by_weekend_weekday_plot <- create_plot(
  data = ride_freq_by_weekend_weekday,
  x_var = "weekday_weekend", y_var ="total_rides", fill_var = "member_casual", 
  plot_type = "bar",
  title = "Ride Frequency (Weekend vs. Weekday)", 
  x_label = "Weekend vs. Weekday", y_label = "Total Rides"
)

print(ride_freq_by_weekend_weekday_plot)
```

#### 

#### **Ride Frequency by Part of Day**

```{r}
# Find the ride frequency by part of day by rider type.
ride_freq_by_day <- cyclistic_data %>% 
  mutate(day_part = factor(
    day_part, levels = c("Morning", "Afternoon", "Evening", "Night"))) %>%
  group_by(member_casual, day_part) %>% 
  summarise(total_rides = n(),
            .groups = "drop")

# Visualize ride frequency by part of day.
ride_freq_by_day_plot <- create_plot(
  data = ride_freq_by_day,
  x_var = "day_part", y_var ="total_rides", fill_var = "member_casual", 
  plot_type = "bar",
  title = "Ride Frequency by Part of Day", 
  x_label = "Part of Day", y_label = "Total Rides"
)

print(ride_freq_by_day_plot)
```

#### Ride Frequency by Hour of Day

```{r}
# Summarize the total rides by hour and rider type
ride_freq_by_hour <- cyclistic_data %>%
  group_by(member_casual, hour = hour(started_at)) %>%
  summarise(total_rides = n(),
            .groups = "drop")

# Create the line chart
ride_freq_by_hour_plot <- create_plot(
  data = ride_freq_by_hour,
  x_var = "hour", y_var ="total_rides", fill_var = "member_casual", 
  plot_type = "line",
  title = "Ride Frequency by Hour of Day", 
  x_label = "Hour", y_label = "Total Rides"
)

print(ride_freq_by_hour_plot)
```

**Ride Frequency During Peak Hours**

```{r}
# Find the peak hour usage by rider type.
peak_hour_stats <- cyclistic_data %>% 
  group_by(member_casual, is_peak_hour) %>% 
  summarise(total_rides = n(), 
            .groups = "drop")

# Visualize ride frequency by day of week.
ggplot(
  peak_hour_stats,
  aes(
    x = is_peak_hour,
    y = total_rides,
    fill = member_casual
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = format_k_m) +  
  scale_fill_manual(values = c("casual" = color_casual, "member" = color_member)) +
  labs(
    title = "Ride Frequency During Peak Hours",
    x = "Is Peak Hour",
    y = "Total Rides",
    fill = "Rider Type"
  ) + 
  theme_minimal()
```

```{r}
# Summarize ride counts for start stations
start_station_popularity <- cyclistic_data %>%
  group_by(start_lat, start_lng, member_casual) %>%
  summarise(start_count = n(), .groups = 'drop')

# Summarize ride counts for end stations
end_station_popularity <- cyclistic_data %>%
  group_by(end_lat, end_lng, member_casual) %>%
  summarise(end_count = n(), .groups = 'drop')

# Assuming the coordinates match exactly for starting and ending points
total_station_popularity <- full_join(start_station_popularity, end_station_popularity, by = c("start_lat" = "end_lat", "start_lng" = "end_lng", "member_casual"))

# Adjusting column names and calculating total popularity
total_station_popularity <- total_station_popularity %>%
  mutate(total_count = start_count + end_count) %>%
  select(lat = start_lat, lng = start_lng, member_casual, total_count)

# Create a map for one user type at a time or both if preferred
map <- leaflet(data = total_station_popularity) %>%
  addTiles() %>%  # Add default OpenStreetMap tile layer
  addHeatmap(lng = ~lng, lat = ~lat, intensity = ~total_count, blur = 20, max = 0.05, radius = 15) %>%
  addLayersControl(
    overlayGroups = c("Casual", "Member"),
    options = layersControlOptions(collapsed = FALSE)
  )

map
```
